Staged Abstract Interpreters

### 2018-11-11 paper/10_intro.tex
section 2

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input 
program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize Futamura projections,
which first identifies the binding-time of variables in the program, i.e., they can be known statically or dynamically, 
and then specializes the program with respect to the given input of program.
However, precisely analyzing binding-time for an arbitray program is hard in general. As an alternative and more practical 
approach to specialization and partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} 
requires the stage annotations (i.e., binding-time annotations) from the programmers, these staging annotations 
control which part of the input program should be specialized.
As a simple example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is statically known, say 5, then we can specialize \texttt{power} with respect to \texttt{e = 5} and 
generate a new specialized function as following:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close connection between 
compilers and interpreters. 
The first Futamura projection specifically shows that specializing an interpreter with respect to the input 
program yields an equivalent executable.
Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed approach to realize Futamura projections,
which first identifies the binding-time of variables in the program, i.e., they can be known statically or dynamically, 
and then specializes the program with respect to the given input of program.
However, precisely analyzing binding-time for an arbitray program is hard in general. As an alternative and programmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} 
requires the stage annotations (i.e., binding-time annotations) from the programmers, these staging annotations 
control which part of the input program should be specialized.
As a simple example, we consider a power function: 
\begin{lstlisting}
def power(b: Int, e: Int): Int = 
  if (e == 0) 1 else e * power(b, e-1)
\end{lstlisting}
If \texttt{e} is statically known, say 5, then we can specialize \texttt{power} with respect to \texttt{e = 5} and 
generate a new specialized function as following:
\begin{lstlisting}
def power5(b: Int) = b * b * b * b * b
\end{lstlisting}

### 2018-11-11 paper/21_bg_interp.tex
section 2

\textbf{Generic Interface.} Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which will be used later for both unstaged/staged and concrete/abstract scenarios.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

@@

\textbf{Stage Polymorphism.}
Readers may notice an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, that is a
type-level function and will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-time purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially being dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

\textbf{Concrete Instantiation.}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for addresses space \texttt{Addr}. The environment and store are just ordinary \texttt{Map} in Scala.
The values can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

\begin{lstlisting}
object ConcInterp extends Concrete {
  type R[+T] = T
  val $\rho$0 = Map[Ident,Addr]();  val $\sigma$0 = Map[Addr,Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = $\sigma$(a)
  def put($\sigma$: Store, a: Addr, v: Value): Store = $\sigma$ + (a -> v)
  def alloc($\sigma$: Store, x: Ident): Addr = $\sigma$.size + 1
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = CloV($\lambda$, $\rho$)
  def num(i: Lit): Value = NumV(i.i)
  def apply_closure(ev: EvalFun)
    (f: Value, arg: Value, $\sigma$: Store): Ans = f match {
      case CloV(Lam(x, e), $\rho$) =>
        val $\alpha$ = alloc($\sigma$, x); val $\rho$_* = put($\rho$, x, $\alpha$)
        val $\sigma$_* = put($\sigma$, $\alpha$, arg); ev(e, $\rho$_*, $\sigma$_*)
    }
  def branch0(cnd: Value, thn: => Ans, els: => Ans): Ans = 
    cnd match { case NumV(i) => if (i == 0) thn else els }
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    (op, v1, v2) match {
      case ('+, NumV(a), NumV(b)) => NumV(a + b)
      ...
    }
}
\end{lstlisting}

>>>>

\paragraph{Generic Interface.} 
Before showing the actual implementation of the concrete semantics, let's first see 
a generic interface for the interpreters, which will be used later for both unstaged/staged and concrete/abstract scenarios.
The trait \texttt{Semantics} first declares several abstract type members, such as \texttt{Addr},
\texttt{Value}, \texttt{Env}, \texttt{Store} and returned type \texttt{Ans}. 
Initial values for environments and stores, and few abstract methods that manipulate environments and stores are also declared.

@@

\paragraph{Stage Polymorphism.}
Readers may notice an abstract higher-kinded type \texttt{R[+T]} is defined in the trait, that is a
type-level function and will be used to specify the binding-time, i.e., staged or not; the annotation 
\texttt{+} indicates type parameter \texttt{T} is covariant.
The LMS framework we use here allows us to specify binding-time purely based on types.
All occurrences of type \texttt{Env}, \texttt{Store}, etc. are wrapped with \texttt{R} 
essentially because they may potentially being dynamic when staging is introduced.
Obviously, the unstaged higher-kinded type then can be simply instantiated as an identity type function: 
\texttt{type R[+T] = T}. In Section~\ref{stagedinterp}, \texttt{R} will be instantiated 
as a staging annotation. \todo{where talk about stage poly?}

\paragraph{Concrete Instantiation.}
Now we can instantiate a concrete interpreter. We first concretize the abstract types as follows.
Since the interpreter is concrete, for the propose that properly models the heap, the type \texttt{Int} 
is used for addresses space \texttt{Addr}. The environment and store are just ordinary \texttt{Map} in Scala.
The values can be either a tagged number \texttt{NumV} or a closure \texttt{CloV} that contains a lambda term
and an environment.

@@

\begin{lstlisting}
object ConcInterp extends Concrete {
  type R[+T] = T
  val $\rho$0 = Map[Ident,Addr]();  val $\sigma$0 = Map[Addr,Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = $\sigma$(a)
  def put($\sigma$: Store, a: Addr, v: Value): Store = $\sigma$ + (a -> v)
  def alloc($\sigma$: Store, x: Ident): Addr = $\sigma$.size + 1
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = CloV($\lambda$, $\rho$)
  def num(i: Lit): Value = NumV(i.i)
  def apply_closure(ev: EvalFun)
    (f: Value, arg: Value, $\sigma$: Store): Ans = f match {
      case CloV(Lam(x, e), $\rho$) =>
        val $\alpha$ = alloc($\sigma$, x); val $\rho$_* = put($\rho$, x, $\alpha$)
        val $\sigma$_* = put($\sigma$, $\alpha$, arg); ev(e, $\rho$_*, $\sigma$_*)
    }
  def branch0(cnd: Value, thn: => Ans, els: => Ans): Ans = 
    cnd match { case NumV(i) => if (i == 0) thn else els }
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    op match {
      case '+ => v1.asInstanceOf[NumV].i + v2.asInstanceOf[NumV].i
      ...
    }
}
\end{lstlisting}

### 2018-11-11 paper/22_bg_abs.tex
section 2

\textbf{Abstract Instantiation.}
Our instantiation roughly follows the Abstracting Abstract Machines methodology 
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the concrete semantics 
to abstract semantics. The store now maps addresses to sets of abstract values, meaning that an address points
to all possible values may occur at runtime, where the abstract value is either a closure or 
a single abstract number element \texttt{NumV}. The address space is also
constrained to be finite, in the monovariant setting, we simply use variable names for it.

@@

\textbf{Stage Polymorphic Lattices.}
To effectively reuse the code between unstaged and staged part, we also make the lattice structure to
be stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

@@

\textbf{Operations Implementation.}

The @get@ and @put@ operation on abstract stores slightly changed according to
the abstract semantics: @get@ uses bottom element in the @Value@ lattice if the queried
address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to a singleton set that
contains one \texttt{CloV} or \texttt{NumV} object.

@@

\textbf{Fixpoint Iteration.}
The interpreter may not terminate
co-inductive caching/truncated depth-first evaluation \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800, Rosendahl:AbsIntPL}

>>>>

\paragraph{Abstract Instantiation.}
Our instantiation roughly follows the Abstracting Abstract Machines methodology 
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the concrete semantics 
to abstract semantics. The store now maps addresses to sets of abstract values, meaning that an address points
to all possible values may occur at runtime, where the abstract value is either a closure or 
a single abstract number \texttt{NumV}, which stands for the top. The address space is also
constrained to be finite, in the monovariant setting, we simply use variable names for it.

@@

\paragraph{Stage Polymorphic Lattices.}
To effectively reuse the code between unstaged and staged part, we also make the lattice structure to
be stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

@@

\paragraph{Operations Implementation.}
The @get@ and @put@ operation on abstract stores slightly changed according to
the abstract semantics: @get@ uses bottom element in the @Value@ lattice if the queried
address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to a singleton set that
contains one \texttt{CloV} or \texttt{NumV} object.

@@

\paragraph{Fixpoint Iteration.}
The interpreter may not terminate
co-inductive caching/truncated depth-first evaluation \cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800, Rosendahl:AbsIntPL}

### 2018-11-11 paper/23_bg_stage.tex
section 2

\textbf{Multi-Stage Programming in LMS.}

@@

Lightweight modular staging (LMS) \todo{cite} is a library-based multi-stage programming framework 
that enables type-safe dynamic code generation. Different from the traditional approach of using
syntactic quasi-quotation, it solely relies on types to distinguish between binding times.

In the example shown above, \texttt{power} is a staged function that takes 2 arguments where 
\texttt{b} is declared as \texttt{Rep[Int]} meaning that \texttt{b} is a representation of type 
\texttt{Int} whose underlying value will be available at the next stage. Meanwhile the actual 
computation result of the \texttt{power} function  will be available at the next stage as well. 
Thus the return value of \texttt{power} is also of type \texttt{Rep[Int]}. The generated code for
\texttt{power5} consists of a function that takes only one argument that corresponds to its 
representation \texttt{b} from the previous stage.

\textbf{Staged Concrete Interpreter.}

We share the same concrete type instantiation, i.e. trait \texttt{Concrete}, with the unstaged version,
but reimplemented the staged version for the concrete operations.

\begin{lstlisting}
trait RepConcInterpOps extends Concrete with LMSOps {
  type R[+T] = Rep[T]
  implicit def valueTyp: Typ[Value]
  val $\rho$0: Rep[Env] = Map[Ident,Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr,Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): Rep[Env] = 
    $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = $\sigma$(a)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): Rep[Store] = 
    $\sigma$ + (a -> v)
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = $\sigma$.size + 1
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = $\lambda$
    val f: Rep[(Value,Store)]=>Rep[(Value,Store)] = 
      (as: Rep[(Value, Store)]) => {
        val arg = as._1; val $\sigma$ = as._2
        val $\alpha$ = alloc($\sigma$, x)
        ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, arg))
      }
    unchecked("CompiledClo(", fun(f), ",", $\lambda$, ",", $\rho$, ")")
  }
  def num(i: Lit): Rep[Value] = {
    unchecked("NumV(", i.i, ")")
  }
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans = {
    val i = unchecked[Int](cnd, ".asInstanceOf[NumV].i")
    (if (i == 0) thn else els).asInstanceOf[Rep[(Value,Store)]] //FIXME: Why?
  }
  def prim_eval(op: Symbol, v1: Rep[Value], v2: Rep[Value]): Rep[Value] = {
    val v1i = unchecked(v1, ".asInstanceOf[NumV].i")
    val v2i = unchecked(v2, ".asInstanceOf[NumV].i")
    unchecked("NumV(", v1i, op.toString.drop(1), v2i, ")")
  }
}
\end{lstlisting}

\textbf{Code Generation.}

code generation, runtime code

>>>>

\paragraph{Multi-Stage Programming in LMS.}
Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a multi-stage programming framework implemented 
as a Scala library that enables dynamic code generation in a type-safe manner.
Different from the syntactic approach such as MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations, 
LMS distinguishes binding-time solely based on types.
LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an expression will be 
known at next stage. All operations act on a @Rep@ expression will be resulted in the generated code. However,
we need to provide intermediate representation and code generation support for type @T@ to LMS. 
Fortunately, the LMS framework already provides
such support for primitive types and commonly used data structure such as arrays and maps; implementing such support
for custom class is also straightforward. Let's go back and see how LMS can be used to specialize the power function
we mentioned in Section 1.

@@

In the code shown above, \texttt{power} takes 2 arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as will, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and body of this function multiples @b@ five times.

\paragraph{Staged Concrete Interpreter.} In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures.

\begin{lstlisting}
trait RepConcInterpOps extends Concrete with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident,Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr,Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): Rep[Env] = 
    $\rho$ + (unit(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = $\sigma$(a)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): Rep[Store] = 
    $\sigma$ + (a -> v)
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = $\sigma$.size + 1
  def num(i: Lit): Rep[Value] = NumV(i)
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans = {
    //FIXME: the cast is ugly, patten match?
    val i = cnd.asInstanceOf[Rep[NumV]].i
    if (i == 0) thn else els
  }
  def prim_eval(op: Symbol, 
    v1: Rep[Value], v2: Rep[Value]): Rep[Value] = op match {
      case '+ => v1.asInstanceOf[Rep[NumV]].i + 
                 v2.asInstanceOf[Rep[NumV]].i
      ...
    }
  ...
}
\end{lstlisting}

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. \todo{NumV}
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, 
however, what we desire is the a \textit{compiled} closure, i.e., the syntactic term of the body expression should be
eliminated and specialized.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which both will be
exhibited in the generated code.

\begin{lstlisting}
def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
  val Lam(x, e) = $\lambda$
  val f: Rep[(Value,Store)]=>Rep[(Value,Store)] = { 
    case (arg: Rep[Value],$\sigma$: Rep[Store]) =>
      val $\alpha$ = alloc($\sigma$, x)
      ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, arg)) 
  }
  CompiledClo(fun(f))
}
\end{lstlisting}

Then, @apply_closure@ takes three arguments the function value, argument value and the latest store, 
which are both of @Rep@ type. It means basically these is nothing we can do at current stage --- 
so we create a new node @ApplyClosure@ that contains these arguments in the intermediate representation graph,
and in the code generation phase we can emit the code that does the application. \todo{reflectEffect}.

\begin{lstlisting}
case class ApplyClosure(f: Rep[Value], 
  arg: Rep[Value], $\sigma$: Rep[Store]) extends Def[(Value, Store)]
def apply_closure(ev: EvalFun)
  (f: Rep[Value], arg: Rep[Value], $\sigma$: Rep[Store]): Ans = {
    reflectEffect(ApplyClosure(f, arg, $\sigma$))
  }
\end{lstlisting}

\paragraph{Code Generation.}
When generating code for the next stage, values like @CompiledClo@ and @NumV@ are kept as what it is because
they are intended to be existed in the next stage; however, we need to treat @ApplyClosure@ a little bit 
differently.

\begin{lstlisting}
  case ApplyClosure(f, arg, $\sigma$) => 
    emitValDef(sym, quote(f) + 
                    ".asInstanceOf[CompiledClo].f(" + 
                    quote(arg) + "," + 
                    quote($\sigma$) + ")")
\end{lstlisting}

Concretely, as shown in the code above, when an @ApplyClosure@ is matched when emitting code, we just 
emit a function application where the function value is extracted from the @CompiledClo@
(we assume the object program is well-typed so that we can safely cast it to @CompiledClo@), 
and argument and store are the rest fields in @ApplyClosure@.
